---
title: "Elizabeth Karan and Trey Johnson Lab 2"
author: "WT Johnson III; Elizabeth Karan"
date: "10/13/2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Exercise: Phylogenetically independent contrasts
```{r}
obj<-read.csv("Centrarchidae.csv",row.names=1); obj
```
Now we can fit a model in which gape.width is affected by buccal.length. Without taking phylogeny into account, these characters do seem to be somewhat weakly correlated:
```{r}
plot(obj[,c("buccal.length","gape.width")])
```
Now let’s fit our OLS regression model to see:
```{r}
fit.ols<-lm(gape.width~buccal.length,data=obj)
fit.ols
summary(fit.ols)
plot(obj[,c("buccal.length","gape.width")])
abline(fit.ols,lwd=2,lty="dashed",col="red")
```
However, we cannot forget that when our data are phylogenetic, the assumption independent and identical distribution of the residual error does not hold. Consequently, we need to take the phylogeny into account. One way to do this is by using PICs:
```{r}
library(ape)
cent.tree<-read.tree("Centrarchidae.tre")
buccal.length<-setNames(obj[,"buccal.length"],rownames(obj))
gape.width<-setNames(obj[,"gape.width"],rownames(obj))
pic.bl<-pic(buccal.length,cent.tree)
pic.gw<-pic(gape.width,cent.tree)
fit.pic<-lm(pic.gw~pic.bl+0)
fit.pic
summary(fit.pic)
plot(pic.bl,pic.gw,xlab="PICs for buccal length",ylab="PICs for gape width")
abline(fit.pic,lwd=2,lty="dashed",col="red")
```
So we can see that in this case, there is not a huge effect of taking phylogeny into consideration. However, it is easy to imagine (& simulate) conditions under which taking the phylogeny into account when fitting a regression model would be of great consequence.
```{r}
library(phytools)

## set seed (so we all get the same result
set.seed(21) ## 10 ok, 2 ok, 3 ok, 6 ok, 7 ok, 

## simulate a coalescent shaped tree
tree<-rcoal(n=100)
plotTree(tree,ftype="off")

## simulate uncorrelated Brownian evolution
x<-fastBM(tree)
y<-fastBM(tree)
par(mar=c(5.1,4.1,2.1,2.1))
plot(x,y)
fit<-lm(y~x)
abline(fit,lwd=2,lty="dashed",col="red")
fit
summary(fit)
anova(fit)

## this is a projection of the tree into morphospace
phylomorphospace(tree,cbind(x,y),label="off",node.size=c(0.5,0.7))
abline(fit,lwd=2,lty="dashed",col="red")
```
```{r}
ix<-pic(x,tree)
iy<-pic(y,tree)
fit<-lm(iy~ix-1) ## we have to fit the model without an intercept term
fit
summary(fit)
anova(fit)
```
Challenge Problem 1: Phylogenetically Independent Contrasts

Fit a OLS (non-phylogenetic) regression model for y~x and a phylogenetic regression model using contrasts. What do you find? Why?

Updated from Liam J. Revell’s Macroevolution in R Course, 2016. Last updated Oct. 2016

Regression Model:
```{r}
picdat<-read.csv("pic-exercise-data.csv",row.names=1);
head(picdat)
picmod<-lm(y~x,data=picdat)
plot(picmod)
title(main="Regression Model",xlab="Y",ylab = "X")
summary(picmod)
anova(picmod)
```
We find that, without using phylogenetic regression, x does not predict y. The p-value is significant for the intercept, but not for the x values.

Phylogenetic Regression Model:
```{r}
library(ape)
pictree<-read.tree("pic-exercise-tree.tre")
picy<-setNames(picdat[,"y"],rownames(picdat))
picx<-setNames(picdat[,"x"],rownames(picdat))
ypic<-pic(picy,pictree)
xpic<-pic(picx,pictree)
picfitting<-lm(xpic~ypic+0)
picfitting
summary(picfitting)
plot(ypic,xpic,xlab="Y Phylo",ylab="X Phylo")
abline(picfitting,lwd=2,lty="dashed",col="red")
```
With phylogenetics encorporated, x strongly predicts y. The p-value is 7.81e-11.

